Raising the Role of Vocabulary Hubs for Semantic Data Interoperability in Dataspaces

### 2024-06-15 EBDVF-2024/paper.qmd
Add files via upload

## GraphDB Graph Database
TODO

@@

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
We aim to go one step further and envision a Semantic Layer for dataspaces, 
which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss 
i) the services provided for the Vocabulary Hub component and 
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

Vocabulary Hub Services:
- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data assets:
- Unstructured data: semantic annotation services for documents based on concept tagging. 
  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data. 
    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
    Ontologies are leveraged to construct the OBDA mapping; 
    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping. 
    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
    and RDF Mapping Language (RML) [@RMLIntroduction2024]
    are intended for data transformation processes within a dataspace, 
    where the original source is used as a basis to create an RDF representation (ETL approach).
  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014]. 
    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
    that allow detailed semantic description of tabular data using a JSON "manifest" file.
  - SKOS concepts are used for structured data in several ways. 
    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
    they can be used to annotate specific literals of an RDF graph, 
    or they can be used to associate documents with (specific parts of) an RDF graph.
- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
    The query can be configured to work for various scenarios.
- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
  - Different versions of the same vocabulary to automatically translate between those versions.
- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
  but intended to represent different levels of abstraction, 
  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.

>>>>

The Vocabulary Hub is one of the core components of any dataspace not only those designed in accordance with IDS-RAM. 
Other reference architecture models, such as GAIA-X architecture document 22.10 Release [@GaiaXArchitectureDocument2022] and Fraunhofer's Reference Architecture Model for the Industrial Data Space [@ReferenceArhictectureModel2017] also describe components with a similar role - the Federated Catalogue and Broker Service, respectively. 
Therefore, our approach designed around the IDS-RAM is adaptable and applicable to data spaces designed in accordance with other reference architecture models.

## GraphDB Graph Database
Ontotext GraphDB is a highly efficient, scalable and robust graph database with RDF and SPARQL support.
It also offers data virtualization capabilities which enable direct access using SPARQL to relational databases, eliminating the need to replicate data.
It can perform semantic inferencing at scale, which enables deriving new semantic facts from existing facts.

GraphDB handles massive loads, queries, and inferencing in real time.
Reasoning and query evaluation are performed over a persistent storage layer. Loading, reasoning, and query evaluation are executed extremely fast even against huge ontologies and knowledge bases.
According to the [LDBC Semantic Publishing Benchmark](https://ldbcouncil.org/benchmarks/spb/), it is one of the most scalable OWL repositories currently available.

@@

## Semantic Layer Approach
The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic building block to achieve semantic interoperability in dataspaces by provisioning common vocabularies. 
Those vocabularies are to be expressed in RDF and by syntactically compatible with OWL to allow for expressive semantics. 
The IDS-RAM does not require, but nonetheless allows for, extending the function of the Vocabulary hub towards providing ontology mappings [@FunctionalLayerIDSRAM2023b], which would enable dataspace connectors to automatically convert data between user-specific and standard data formats. Such automatic conversions are to be performed by Data Tranformation Apps by either the Data Consumers, the Data Providers, or both.

We aim to achieve this through a Semantic Layer for dataspaces, 
which provides vocabulary-based services for advanced and automated semantic descriptions of metadata and data. 
The Semantic Layer is implemented as services based on the GraphDB and PoolParty and provided as a one seamlessly integrated component.

In the following, we discuss

i) the services provided for the Vocabulary Hub component and 
ii) additional services as Vocabulary Hub extensions that implement a Semantic Layer for dataspaces.

Vocabulary Hub Services:

- RDF based: both GraphDB and PoolParty store and manage data based on the RDF data model. RDF is the basic building stone and underlying vocabulary definitions.
- Scalable vocabulary storage: cluster implementation for higher throughput and a fault tolerant architecture.
- Knowledge graph management: easy management of vocabularies using a rich web user interface and integration via Web APIs. 
  Vocabularies can be created and edited based on standards like RDFS, OWL and SKOS, and existing vocabularies can be reused and combined to fit specific dataspace needs.
- SPARQL service for semantic lookups: dereference URIs describing data assets to retrieve a semantic description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data assets:

- Unstructured data: semantic annotation services for documents based on concept tagging. 
  PoolParty’s KG-based NLP primarily uses SKOS taxonomies to identify concepts in documents, 
  but can be extended with OWL ontology elements (classes, properties) for increased expressiveness.
- Structured data: we support structured data harmonization methods, including semantic mapping and linking SKOS concepts
  - Ontology-Based Data Access (OBDA) is used for virtualization of relational data, i.e. accessing relational data as semantic data. 
    It is well suited for semantic querying of large amounts of data, while keeping the original relational data in place (NoETL approach). 
    Ontologies are leveraged to construct the OBDA mapping; 
    they are governed via the Semantic Layer and can harmonize data assets of various formats using a precise mapping [@andreselAdaptingOntologybasedData2024]. 
    Furthermore, it can be used to establish a fine-grained access control, thereby deciding for individual data elements if they should be exposed ot not.
  - RDB to RDF Mapping Language (R2RML) [@souripriyadasR2RMLRDBRDF2012]
    and RDF Mapping Language (RML) [@RMLIntroduction2024]
    are intended for data transformation processes within a dataspace, 
    where the original source is used as a basis to create an RDF representation (ETL approach).
  - GraphDB supports both OBDA and R2RML through its integration of ONTOP [@bagosiOntopFrameworkOntology2014]. 
    The same mappings can be used for NoETL virtual data access, or for ETL (to materialize the semantic data).
  - CSV on the Web (CSVW) [@swirrlCSVWStandards2024] is a set of W3C standards 
    that allow detailed semantic description of tabular data using a JSON "manifest" file.
  - SKOS concepts are used for structured data in several ways. 
    Concepts can be directly added to RDF data via semantic links (properties from an ontology), 
    they can be used to annotate specific literals of an RDF graph, 
    or they can be used to associate documents with (specific parts of) an RDF graph.
- Inference services: we provide metadata expansion services based on interlinked vocabularies in combination with ontologies. 
  This inference approach can be used with semantic tagging for unstructured data, as well as the structured data approaches.
  - Inference tagging is provided by a PoolParty tagging service in combination with an expansion query for implementing the inference. 
    The query can be configured to work for various scenarios.
- Vocabulary crosswalks are a data modeling approach using interlinks to relate:
  - Different vocabularies from similar or the same domain that cover different areas or aspects of the domain. 
    By relating concepts with each other using 1:1, 1:n and m:n relations, we can automatically expand queries and annotations by traversing the interlinked graph.
  - Different versions of the same vocabulary to automatically translate between those versions.
- Metadata inference based on interlinked vocabularies is similar to crosswalks, 
  but intended to represent different levels of abstraction, 
  so that we are able to infer from a more general vocabulary to a more specific vocabulary and vice versa. 
  For example, we can start with DCAT themes vocabularies and link to further vocabularies, which are more specific for certain areas or which go into more detail.

### 2024-06-15 EBDVF-2024/paper.tex
Add files via upload

@@

The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic
building block to achieve semantic interoperability in dataspaces by
provisioning common vocabularies. We aim to go one step further and
envision a Semantic Layer for dataspaces, which provides
vocabulary-based services for advanced and automated semantic
descriptions of metadata and data. The Semantic Layer is implemented as
services based on the GraphDB and PoolParty and provided as a one
seamlessly integrated component.

In the following, we discuss i) the services provided for the Vocabulary
Hub component and ii) additional services as Vocabulary Hub extensions
that implement a Semantic Layer for dataspaces.

Vocabulary Hub Services: - RDF based: both GraphDB and PoolParty store
and manage data based on the RDF data model. RDF is the basic building
stone and underlying vocabulary definitions. - Scalable vocabulary
storage: cluster implementation for higher throughput and a fault
tolerant architecture. - Knowledge graph management: easy management of
vocabularies using a rich web user interface and integration via Web
APIs. Vocabularies can be created and edited based on standards like
RDFS, OWL and SKOS, and existing vocabularies can be reused and combined
to fit specific dataspace needs. - SPARQL service for semantic lookups:
dereference URIs describing data assets to retrieve a semantic
description of the entity from the graph database.

Semantic Layer Services support both unstructured and structured data
assets: - Unstructured data: semantic annotation services for documents
based on concept tagging. PoolParty's KG-based NLP primarily uses SKOS
taxonomies to identify concepts in documents, but can be extended with
OWL ontology elements (classes, properties) for increased
expressiveness. - Structured data: we support structured data
harmonization methods, including semantic mapping and linking SKOS
concepts - Ontology-Based Data Access (OBDA) is used for virtualization
of relational data, i.e.~accessing relational data as semantic data. It
is well suited for semantic querying of large amounts of data, while
keeping the original relational data in place (NoETL approach).
Ontologies are leveraged to construct the OBDA mapping; they are
governed via the Semantic Layer and can harmonize data assets of various
formats using a precise mapping. Furthermore, it can be used to
establish a fine-grained access control, thereby deciding for individual
data elements if they should be exposed ot not. - RDB to RDF Mapping
Language (R2RML) \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping
Language (RML) \citep{RMLIntroduction2024} are intended for data
transformation processes within a dataspace, where the original source
is used as a basis to create an RDF representation (ETL approach). -
GraphDB supports both OBDA and R2RML through its integration of ONTOP
\citep{bagosiOntopFrameworkOntology2014}. The same mappings can be used
for NoETL virtual data access, or for ETL (to materialize the semantic
data). - CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set
of W3C standards that allow detailed semantic description of tabular
data using a JSON ``manifest'' file. - SKOS concepts are used for
structured data in several ways. Concepts can be directly added to RDF
data via semantic links (properties from an ontology), they can be used
to annotate specific literals of an RDF graph, or they can be used to
associate documents with (specific parts of) an RDF graph. - Inference
services: we provide metadata expansion services based on interlinked
vocabularies in combination with ontologies. This inference approach can
be used with semantic tagging for unstructured data, as well as the
structured data approaches. - Inference tagging is provided by a
PoolParty tagging service in combination with an expansion query for
implementing the inference. The query can be configured to work for
various scenarios. - Vocabulary crosswalks are a data modeling approach
using interlinks to relate: - Different vocabularies from similar or the
same domain that cover different areas or aspects of the domain. By
relating concepts with each other using 1:1, 1:n and m:n relations, we
can automatically expand queries and annotations by traversing the
interlinked graph. - Different versions of the same vocabulary to
automatically translate between those versions. - Metadata inference
based on interlinked vocabularies is similar to crosswalks, but intended
to represent different levels of abstraction, so that we are able to
infer from a more general vocabulary to a more specific vocabulary and
vice versa. For example, we can start with DCAT themes vocabularies and
link to further vocabularies, which are more specific for certain areas
or which go into more detail.

>>>>

The Vocabulary Hub is one of the core components of any dataspace not
only those designed in accordance with IDS-RAM. Other reference
architecture models, such as GAIA-X architecture document 22.10 Release
\citep{GaiaXArchitectureDocument2022} and Fraunhofer's Reference
Architecture Model for the Industrial Data Space
\citep{ReferenceArhictectureModel2017} also describe components with a
similar role - the Federated Catalogue and Broker Service, respectively.
Therefore, our approach designed around the IDS-RAM is adaptable and
applicable to data spaces designed in accordance with other reference
architecture models.

@@

The Vocabulary Hub, as IDS defines it in the IDS-RAM, is a basic
building block to achieve semantic interoperability in dataspaces by
provisioning common vocabularies. Those vocabularies are to be expressed
in RDF and by syntactically compatible with OWL to allow for expressive
semantics. The IDS-RAM does not require, but nonetheless allows for,
extending the function of the Vocabulary hub towards providing ontology
mappings \citep{FunctionalLayerIDSRAM2023b}, which would enable
dataspace connectors to automatically convert data between user-specific
and standard data formats. Such automatic conversions are to be
performed by Data Tranformation Apps by either the Data Consumers, the
Data Providers, or both.

We aim to achieve this through a Semantic Layer for dataspaces, which
provides vocabulary-based services for advanced and automated semantic
descriptions of metadata and data. The Semantic Layer is implemented as
services based on the GraphDB and PoolParty and provided as a one
seamlessly integrated component.

In the following, we discuss

\begin{enumerate}
\def\labelenumi{\roman{enumi})}
\tightlist
\item
  the services provided for the Vocabulary Hub component and
\item
  additional services as Vocabulary Hub extensions that implement a
  Semantic Layer for dataspaces.
\end{enumerate}

Vocabulary Hub Services:

\begin{itemize}
\tightlist
\item
  RDF based: both GraphDB and PoolParty store and manage data based on
  the RDF data model. RDF is the basic building stone and underlying
  vocabulary definitions.
\item
  Scalable vocabulary storage: cluster implementation for higher
  throughput and a fault tolerant architecture.
\item
  Knowledge graph management: easy management of vocabularies using a
  rich web user interface and integration via Web APIs. Vocabularies can
  be created and edited based on standards like RDFS, OWL and SKOS, and
  existing vocabularies can be reused and combined to fit specific
  dataspace needs.
\item
  SPARQL service for semantic lookups: dereference URIs describing data
  assets to retrieve a semantic description of the entity from the graph
  database.
\end{itemize}

Semantic Layer Services support both unstructured and structured data
assets:

\begin{itemize}
\tightlist
\item
  Unstructured data: semantic annotation services for documents based on
  concept tagging. PoolParty's KG-based NLP primarily uses SKOS
  taxonomies to identify concepts in documents, but can be extended with
  OWL ontology elements (classes, properties) for increased
  expressiveness.
\item
  Structured data: we support structured data harmonization methods,
  including semantic mapping and linking SKOS concepts

  \begin{itemize}
  \tightlist
  \item
    Ontology-Based Data Access (OBDA) is used for virtualization of
    relational data, i.e.~accessing relational data as semantic data. It
    is well suited for semantic querying of large amounts of data, while
    keeping the original relational data in place (NoETL approach).
    Ontologies are leveraged to construct the OBDA mapping; they are
    governed via the Semantic Layer and can harmonize data assets of
    various formats using a precise mapping
    \citep{andreselAdaptingOntologybasedData2024}. Furthermore, it can
    be used to establish a fine-grained access control, thereby deciding
    for individual data elements if they should be exposed ot not.
  \item
    RDB to RDF Mapping Language (R2RML)
    \citep{souripriyadasR2RMLRDBRDF2012} and RDF Mapping Language (RML)
    \citep{RMLIntroduction2024} are intended for data transformation
    processes within a dataspace, where the original source is used as a
    basis to create an RDF representation (ETL approach).
  \item
    GraphDB supports both OBDA and R2RML through its integration of
    ONTOP \citep{bagosiOntopFrameworkOntology2014}. The same mappings
    can be used for NoETL virtual data access, or for ETL (to
    materialize the semantic data).
  \item
    CSV on the Web (CSVW) \citep{swirrlCSVWStandards2024} is a set of
    W3C standards that allow detailed semantic description of tabular
    data using a JSON ``manifest'' file.
  \item
    SKOS concepts are used for structured data in several ways. Concepts
    can be directly added to RDF data via semantic links (properties
    from an ontology), they can be used to annotate specific literals of
    an RDF graph, or they can be used to associate documents with
    (specific parts of) an RDF graph.
  \end{itemize}
\item
  Inference services: we provide metadata expansion services based on
  interlinked vocabularies in combination with ontologies. This
  inference approach can be used with semantic tagging for unstructured
  data, as well as the structured data approaches.

  \begin{itemize}
  \tightlist
  \item
    Inference tagging is provided by a PoolParty tagging service in
    combination with an expansion query for implementing the inference.
    The query can be configured to work for various scenarios.
  \end{itemize}
\item
  Vocabulary crosswalks are a data modeling approach using interlinks to
  relate:

  \begin{itemize}
  \tightlist
  \item
    Different vocabularies from similar or the same domain that cover
    different areas or aspects of the domain. By relating concepts with
    each other using 1:1, 1:n and m:n relations, we can automatically
    expand queries and annotations by traversing the interlinked graph.
  \item
    Different versions of the same vocabulary to automatically translate
    between those versions.
  \end{itemize}
\item
  Metadata inference based on interlinked vocabularies is similar to
  crosswalks, but intended to represent different levels of abstraction,
  so that we are able to infer from a more general vocabulary to a more
  specific vocabulary and vice versa. For example, we can start with
  DCAT themes vocabularies and link to further vocabularies, which are
  more specific for certain areas or which go into more detail.
\end{itemize}
